---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r load_process_data, echo = TRUE}

data_file <- "activity.zip"
activity <- read.csv(unz(data_file, "activity.csv"), colClasses = c("date" = "Date"))

```


## What is mean total number of steps taken per day?

```{r mean_steps, echo = TRUE}

steps_per_day <- aggregate(steps ~ date, data = activity, FUN = sum, na.rm = TRUE)
hist(steps_per_day$steps, main ="Histogram of Steps per Day", xlab = "Steps per Day")
mn <- as.character(round(mean(steps_per_day$steps),2))
mdn <- as.character(round(median(steps_per_day$steps),2))

```

The mean number of steps per day is `r mn` and the median is `r mdn`.

## What is the average daily activity pattern?

```{r daily_averages, echo = TRUE}

avg_per_interval <- aggregate(steps ~ interval, data = activity, FUN = mean, na.rm = TRUE)
plot(avg_per_interval$steps, type="l", xlab="Interval", ylab="Average No. of Steps")
max <- as.character(which.max(avg_per_interval$steps))
hour = as.character(round(as.integer(max)/12,2))

```

The interval that contains the highest number of steps on average is `r max` which is around `r hour` hours into the day.

## Imputing missing values

```{r fill_missing_values, echo=TRUE}

na_row_count = as.character(nrow(activity[which(is.na(activity$steps)),]))

# Take a copy of activity data frame
filled_activity = activity

# Fill in NA values of data frame with averages for that interval.
for(i in 1:nrow(filled_activity)){
    if(is.na(filled_activity[i,1])){
        interval = filled_activity[i, 3]
        average_for_interval = avg_per_interval$steps[match(interval, avg_per_interval$interval)]
        filled_activity[i,1] <- average_for_interval
    } 
}

# Recalculate Steps Per Day
steps_per_day <- aggregate(steps ~ date, data = activity, FUN = sum, na.rm = TRUE)
hist(steps_per_day$steps, main ="Histogram of Steps per Day", xlab = "Steps per Day")
mn2 <- as.character(round(mean(steps_per_day$steps),2))
mdn2 <- as.character(round(median(steps_per_day$steps),2))
```

The number of rows with NAs was `r na_row_count`. These have been replaced with the average for the missing time interval.

The mean number of steps per day is `r mn2` and the median is `r mdn2`. Compared with mean of `r mn` and median of `r mdn` from the original data set.


## Are there differences in activity patterns between weekdays and weekends?

```{r weekend, echo=TRUE}

# Create new 'week' vector
weekend = c()
for(i in 1:nrow(filled_activity)){
    if(weekdays(filled_activity[i,2]) %in% c("Saturday", "Sunday")){
        weekend[i] <- "weekend"
    } else{
        weekend[i] <- "weekday"
    }
}
# Add 'week' vector to activity data frame as a factor
filled_activity$week <- as.factor(weekend)
avg_per_interval_week = aggregate(steps ~ interval + week, data = filled_activity, FUN = mean)

# Create Panel Plot with Lattice
library(lattice)
xyplot(steps ~ interval | week, data = avg_per_interval_week, type = "l",  layout = c(1,2))
```