---  
title: Statistics Notes
output: 
  html_document:
    toc: true
    theme: null
    css: "css/style.css"
---

Last Updated: December 2015

To Do:

* Regression

Course notes for Statistics.

Notes are mainly for the [Data Science Specialisation](https://www.coursera.org/specializations/jhudatascience) on Coursera. See references.Rmd for more sources.

## Intro to Statistics

Anecdotal evidence is misleading because: small no. of observations, selection bias, confirmation bias, inaccuracy. 
A statistical approach tries to avoid these problems. It involves data collection, descriptive of stats, exploratory data analysis, inference and hypothesis testing and estimation.

A _population_ encompasses everything or everyone that you want to make statements about. A study will usually be carried out on a smaller subset or _sample_ of the population. A group of respondents is called a _cohort_.

The cohort is supposed to be representative of the population (but sometime studies are weighted towards some section of the population by design or by necessity).

A _cross-sectional_ study captures a snapshot of a group at a point in time, where a _longitudinal_ study follows a group repeatedly over time.

A cross-sectional study captures a snapshot of a group at a point in time, where a longitudinal study follows a group repeatedly over time.

Statistics can be used to determine whether the data collected 'means something'. 

* It tries to be precise about the data using summary statistics like mean, median and variance. 
* It accounts for the possibility of the results occuring randomly, assessing _statistical significance_ of the result. 
* And it tries to assess the effect of errors or bias on the data.

```{r child = '1descriptive_stats.Rmd'}
```
```{r child = '2probability.Rmd'}
```
```{r child = '3testing.Rmd'}
```
```{r child = '4resampling.Rmd'}
```

# Appendices

```{r child = 'zglossary.Rmd'}
```

```{r child = 'zreferences.Rmd'}
```
