---
output: html_document
---
# USA: Harmful Weather Events

## Synopsis

Analysis of Storm Data from the U.S. National Oceanic and Atmospheric Administration to answer the following questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

## Requirements

The _stringdist_ package is required.

```{r packages, echo = TRUE}
library(stringdist)
```


## Data Processing

Load data from the NOAA Storm Database, available from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2.

```{r load_process_data, echo = TRUE}

# Load Data from file
if(!exists("src_data")){
    data_file = "data/repdata_data_StormData.csv"
    src_data <- read.csv(data_file, stringsAsFactors = FALSE)
}

# Subset to save memory
# src_data <- subset(src_data[190000:200000,c(2, 8, 23, 24, 25, 26,27, 28)], CROPDMG > 0)
src_data <- subset(src_data[,c(2, 8, 23, 24, 25, 26,27, 28)], CROPDMG > 0)
```

To calculate the most costly events the event type (EVTYPE) field needs to be cleaned up and mapped to the clean list of events defined in _National Weather Service Instruction 10-1605_.

This is done by comparing the EVTYPE value with the events listed and choosing the closest match.

<table>
<tbody>
<tr class="firstRow">
<td class="val0-cell">Astronomical Low Tide </td>
<td class="val1-cell">Frost/Freeze </td>
<td class="val2-cell">Marine High Wind </td>
<td class="val3-cell">Wildfire </td>
</tr>
<tr>
<td class="val0-cell">Avalanche </td>
<td class="val1-cell">Funnel Cloud </td>
<td class="val2-cell">Marine Strong Wind</td>
<td class="val3-cell">Winter Storm </td>
</tr>
<tr>
<td class="val0-cell">Blizzard </td>
<td class="val1-cell">Freezing Fog</td>
<td class="val2-cell">Marine Thunderstorm Wind </td>
<td class="val3-cell">Winter Weather </td>
</tr>
<tr>
<td class="val0-cell">Coastal Flood </td>
<td class="val1-cell">Hail</td>
<td class="val2-cell">Rip Current </td>
<td class="val3-cell"></td>
</tr>
<tr>
<td class="val0-cell">Cold/Wind Chill</td>
<td class="val1-cell">Heat </td>
<td class="val2-cell">Seiche</td>
<td class="val3-cell"></td>
</tr>
<tr>
<td class="val0-cell">Debris Flow </td>
<td class="val1-cell">Heavy Rain </td>
<td class="val2-cell">Sleet </td>
<td class="val3-cell"></td>
</tr>
<tr>
<td class="val0-cell">Dense Fog </td>
<td class="val1-cell">Heavy Snow </td>
<td class="val2-cell">Storm Surge/Tide</td>
<td class="val3-cell"></td>
</tr>
<tr>
<td class="val0-cell">Dense Smoke </td>
<td class="val1-cell">High Surf </td>
<td class="val2-cell">Strong Wind </td>
<td class="val3-cell"></td>
</tr>
<tr>
<td class="val0-cell">Drought </td>
<td class="val1-cell">High Wind </td>
<td class="val2-cell">Thunderstorm Wind</td>
<td class="val3-cell"></td>
</tr>
<tr>
<td class="val0-cell">Dust Devil </td>
<td class="val1-cell">Hurricane (Typhoon) </td>
<td class="val2-cell">Tornado </td>
<td class="val3-cell"></td>
</tr>
<tr>
<td class="val0-cell">Dust Storm </td>
<td class="val1-cell">Ice Storm </td>
<td class="val2-cell">Tropical Depression</td>
<td class="val3-cell"></td>
</tr>
<tr>
<td class="val0-cell">Excessive Heat</td>
<td class="val1-cell">Lake-Effect Snow</td>
<td class="val2-cell">Tropical Storm </td>
<td class="val3-cell"></td>
</tr>
<tr>
<td class="val0-cell">Extreme Cold/Wind Chill</td>
<td class="val1-cell">Lakeshore Flood</td>
<td class="val2-cell">Tsunami </td>
<td class="val3-cell"></td>
</tr>
<tr>
<td class="val0-cell">Flash Flood</td>
<td class="val1-cell">Lightning </td>
<td class="val2-cell">Volcanic Ash </td>
<td class="val3-cell"></td>
</tr>
<tr class="lastRow">
<td class="val0-cell">Flood</td>
<td class="val1-cell">Marine Hail </td>
<td class="val2-cell">Waterspout </td>
<td class="val3-cell"></td>
</tr>
</tbody>
</table>

The PROPDMGEXP and CROPDMGEXP fields are also cleaned up to provide consistent multipliers.

```{r clean_up, echo = TRUE}
# List of Events
events=c("Astronomical Low Tide","Avalanche","Blizzard","Coastal Flood","Cold/Wind Chill","Debris Flow","Dense Fog","Dense Smoke","Drought","Dust Devil","Dust Storm","Excessive Heat","Extreme Cold/Wind Chill","Flash Flood","Flood","Frost/Freeze","Funnel Cloud","Freezing Fog","Hail","Heat","Heavy Rain","Heavy Snow","High Surf","High Wind","Hurricane (Typhoon)","Ice Storm","Lake-Effect Snow","Lakeshore Flood","Lightning","Marine Hail","Marine High Wind","Marine Strong Wind","Marine Thunderstorm Wind","Rip Current","Seiche","Sleet","Storm Surge/Tide","Strong Wind","Thunderstorm Wind","Tornado","Tropical Depression","Tropical Storm","Tsunami","Volcanic Ash","Waterspout","Wildfire","Winter Storm","Winter Weather")

for(i in 1:nrow(src_data)){
    
    # Map Event Types (EVTYPE) to the nearest event from list above.
    src_data$event[i] <- events[amatch(src_data$EVTYPE[i], events, maxDist=Inf)]
    
    # Clean up Prop Damage Multipliers
    if(src_data$PROPDMGEXP[i] %in% c("", "-", "?")) {src_data$PROPDMGEXP[i] = "Z"}
    else if (src_data$PROPDMGEXP[i] %in% c(0:8)){src_data$PROPDMGEXP[i] = "T"}
    else if(src_data$PROPDMGEXP[i] == "+"){src_data$PROPDMGEXP[i] = "O"}
    else {src_data$PROPDMGEXP[i] = toupper(src_data$PROPDMGEXP[i])}
    
    # Clean up Crop Damage Multipliers
    if(src_data$CROPDMGEXP[i] %in% c("", "-", "?")) {src_data$CROPDMGEXP[i] = "Z"}
    else if (src_data$CROPDMGEXP[i] %in% c(0:8)){src_data$CROPDMGEXP[i] = "T"}
    else if(src_data$CROPDMGEXP[i] == "+"){src_data$CROPDMGEXP[i] = "O"}
    else {src_data$CROPDMGEXP[i] = toupper(src_data$CROPDMGEXP[i])}
}
```

Now the impact of each weather event can be calculated. The health impact has been calculated as the sum of the number of injuries plus the number of fatalities.

The Propert and Crop damage has been multiplied by the appropriate amount to give a consistent total monetary value.

```{r calculate_impact, echo=TRUE}

# Calculate Impact on Population Health
src_data$health_impact = src_data$FATALITIES + src_data$INJURIES

# Calculate Impact on Property
multipliers = c("Z"=0, "O"=1, "T"=10, "H" = 100, "K"=100, "M"=1000000, "B"=1000000000)
src_data$property_impact = src_data$PROPDMG*multipliers[as.character(src_data$PROPDMGEXP)]
src_data$crop_impact = src_data$CROPDMG*multipliers[src_data$CROPDMGEXP]
```

## Results

The health, property and crop impact is calculated as a total per event. The top 5 events are summarised and displayed below.

```{r calculate_results, echo=TRUE}

data_summ <- aggregate(cbind(health_impact, property_impact, crop_impact) ~ event, data = src_data, FUN = sum)

health <- data_summ[, 1:2]
health <- health[order(-health$health_impact),]
health <- health[1:5,]
barplot(health$health_impact, names.arg = health$event, main="Top 5 Events (Impact on health)", horiz=TRUE, las="2", cex.names="0.75", cex.axis="0.75")
health

property <- data_summ[, c(1,3)]
property <- property[order(-property$property_impact),]
property <- property[1:5,]
property$property_impact <- property$property_impact/1000000
barplot(property$property_impact, names.arg = property$event, main="Top 5 Events (Impact on property)", horiz=TRUE, xlab="millions", las="2", cex.names="0.75", cex.axis="0.75")
property

crop <- data_summ[, c(1,4)]
crop <- crop[order(-crop$crop_impact),]
crop <- crop[1:5,]
crop$crop_impact <- crop$crop_impact/1000000
barplot(crop$crop_impact, names.arg = crop$event, main="Top 5 Events (Impact on crop)", horiz=TRUE, xlab="millions", las="2", cex.names="0.75", cex.axis="0.75")
crop
```

